SMT-Based Generation of Symbolic Automata \thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2019-03-31 2017-Xudong-SMT/ActaInformatica.tex
cosmetic changes

Implementing these semantics raised several challenges, in order:
\begin{itemize}
	\item to obtain a tool that could be applied to pNets representing
	various languages, in particular various action algebras,
	with their specific decision theories;
	\item to separate clearly the 
	algorithm generating the transitions of the open automaton from
	combination of all possible (symbolic) behaviours, and from
	the symbolic reasoning part, specifically here using an SMT
	engine to check the
	satisfiability of the predicates generated by our algorithm;
	\item to build a prototype and validate the approach on our basic
	case-studies, and understand the efficiency of the interaction
	with the SMT solver.
\end{itemize}

@@

\paragraph{Contribution}
In the current work we contribute in the following aspects.
\begin{itemize}
	\item We present an open automaton generation algorithm. We
	implement a full working prototype, within the 
	VerCors platform. In the process, we improve the
	semantic rules from~\cite{henrio:Forte2016}, and add features in
	the algorithm to deal with the full 
	model, including management of variables and assignments.
	\item We implement the interaction between our algorithm and the Z3
	SMT solver, for checking satisfiability of the transitions
	generated by the algorithm.
	\item We define a strategy checking satisfiability of open transitions in the course of the residual algorithm, and show that  the reachable part of the open automaton can be generated and it is minimal with respect to the decidability of predicate satisfiability.
	\item We show the usage of this approach on various use-cases from different settings, including an
	industry-inspired case-study, namely one architectural pattern
	extracted (and extended) from the BIP specification of a nanosatellite on-board
	software.
\end{itemize}

@@

\paragraph{Structure.}
In Section
\ref{section:pnets} we give a description and a formal definition of
the pNet model. 
In Section \ref{section:op-semantics} we briefly recall  the operational semantics
of pNets.
In Section \ref{section:implementation} we present an algorithm to compute this semantics, including the interaction with Z3.
In Section~\ref{section:use-cases} we give a few use-cases, in particular, a BIP
architecture from the nano-satellite case-study. 
Finally we conclude and discuss perspectives in Section
\ref{section:conclusion}.

@@

\renewcommand{\P}{\mathcal V}
\def\Talg{\mathcal{T}_{\Sigma,\P}}
Formally, we assume the existence of a term algebra $\Talg$,
where $\Sigma$ is the signature of the data and action constructors,
and $\P$ a set of variables. Within $\Talg$, we distinguish a set of
data expressions $\mathcal{E}_\P$, including a set of Boolean
expressions $\mathcal{B}_{\P}$ ($\mathcal{B}_{\P}\subseteq\mathcal{E}_\P$).
On top of $\mathcal{E}_\P$ we build the action algebra
$\mathcal{A}_\P$, with $\mathcal{A}_\P\subseteq\mathcal{T}_\P,
\mathcal{E}_\P\cap\mathcal{A}_\P=\emptyset$;
naturally action terms will use data expressions as sub-terms.
The function $\vars(t)$ identifies the set of variables in a term
$t\in\AlgT$.

@@

\begin{definition}[PLTS]
	\label{PLTS}
	%\noteSB{It's a bit strange that algebra presentations are not referred to here}
	Given a term algebra $\Talg$, a PLTS is a tuple
	$PLTS\triangleq\mylangle S,s_0, \to\myrangle$ where:
	\begin{itemize}
		\item[$\bullet$]
		$S$ is a set of \emph{states}, with $s_0 \in S$ the \emph{initial state}.
		\item[$\bullet$] $\to \subseteq S \times L \times S$ is the \emph{transition relation}, with
		$L$ the set of labels of the form
		$\langle \alpha,~e_b,~(x_j\!:= {e}_j)^{j\in J}\rangle$,
		where $\alpha \in\AlgA_{\P}$ is a parameterised action,
		$e_b \in \AlgB_{\P}$ is a guard, and
		expressions  $\AlgE_{\P}\cup\AlgA_{P}$ are assigned to $x_j$.
		If 
		$s \xrightarrow{\langle \alpha,~e_b,~(x_j\!:= {e}_j)^{j\in
				J}\rangle} s'$ then 
		$\vars(e_b)\!\subseteq\! \vars(s)\cup\vars(\alpha)$, and
		$\forall j\!\in\! J .\,\vars(e_j)\!\subseteq\! \vars(s)\land 
		x_j\!\in\!\vars(s')$.
		%\item[$\bullet$]
		%The $x_j$ are \emph{state variables} of %% the state $s'$. 
	\end{itemize}
\end{definition}

@@

%% \begin{definition}[pNets]\label{def-pnets}
%% A pNet is a hierarchical structure, whereof the leaves are PLTSs and holes:
%% $\pNet::= PLTS~|~\mylangle \pNet_i^{i\in I}, S_j^{j\in J}, \symb{SV}_k^{k\in K}\myrangle$,
%% where
%% \begin{itemize}
%% \item[$\bullet$] %\noteEM{Pire que ca, $\I$ n'est pas
%%                  %defini...}\noteSB{Why is one $\I$ and the other
%%   %$\I_\P$?}
%%   $I$ and $J$ are the sets over which are indexed sub-pNets and holes, respectively, such that: $I\cap J=\emptyset$ and  $I\cup J\neq\emptyset$.

@@

\begin{definition}[pNets]\label{def-pnets}
	A pNet is a hierarchical structure where leaves are PLTSs and holes:
	%%  \footnote{Typing can 
	%% be added to pNets by defining and checking their interfaces, called 
	%% sort~\cite{HMZ-FORTE2016}}
	$\pNet\triangleq PLTS~|~\mylangle \pNet_i^{i\in I}, J, \symb{SV}_k^{k\in 
		K}\myrangle$, with $I, J, K$ potentialy infinite,
	where
	\begin{itemize}
		\item[$\bullet$] $\pNet_i^{i\in I}$ is the family of sub-pNets;
		%  $\pNet_i^{i\in I}$ is a family of sub-pNets where $I\in\I_\P$ is the set over which 
		%sub-pNets are indexed.

		\item[$\bullet$] $J$ is a set of indexes, called \emph{holes}.
		$I$ and $J$ are \emph{disjoint}: $I\!\cap\! J=\emptyset$,  $I\!\cup\! J\neq\emptyset$
		%\item[$\bullet$] $\Sort_j \subseteq \AlgAS$ is a set of action terms, denoting the 
		%\emph{sort} of
		%hole $j$.

		\item[$\bullet$] $\symb{SV}_k^{k\in K}$ is a set of
		synchronisation vectors ($K\in\I_\P$). $\forall k\!\in\! K,
		\symb{SV}_k\!=\!\alpha_{l}^{l\in I_k \uplus J_k}\to\alpha'_k\,|\,g_k$, where
		$\alpha'_k\in \actions{\variables}$, $I_k\subseteq I$, $J_k\subseteq J$, and 
		$\vars(\alpha'_k)\subseteq \bigcup_{l\in I_k\uplus 
			J_k}{\vars({\alpha_l})}$. The global action of a vector $\symb{SV}_k$ is
		$\alpha'_k$. The Boolean expression $g_k $, such that $\vars(g_k)\subseteq \bigcup_{l\in 
			I_k\uplus J_k}{\vars({\alpha_l})}$, is a guard associated to the vector.

@@

  \includegraphics[width=\linewidth]{ActaXFIG/Enable1}
  \\[1.3ex]
 \includegraphics[width=\linewidth]{ActaXFIG/Enable2}
  \caption{Two pNet encodings for  Enable }  \label{schema:enable-pnets}
\end{minipage}
  \hspace{2mm}
\begin{minipage}{6cm}
  \includegraphics[width=\linewidth]{ActaXFIG/P-QR}
  \caption{Composed pNet for ``P>>(Q>>R)''}  \label{schema:enable-composed}
\end{minipage}

@@

	To give simple intuitions of the open pNet model and its semantics, we use
	here a simple example based on the ``Enable'' operator of the LOTOS language.
	In the Enable expression ``P>>Q'', an \texttt{exit} statement within P
	terminates the current process, carrying a value $x$ that
	is captured by the \texttt{accept} statement of Q. 	
	We use a simple action algebra,
	containing two constructors $\delta(x)$ and $acc(x)$, for any
	possible data type of the variable $x$, corresponding to the
	statements \texttt{exit} and \texttt{accept}. 
	We need no specific predicate over the action expressions, apart
	from equality of actions.

@@

	Note that synchronisation vectors are defined in a parameterised
	manner:  the first and third lines represent one vector
	for each possible action of hole $P$
	(resp. $Q$). This notation can also use predicates, as in the first
	case, in which we want the vector to apply to any action of $P$
	except $\delta(z)$.

	In Figure \ref{schema:enable-composed}, we enrich our example by
	composing two occurences of the EnableState pNet. In Section
        \ref{section:use-case:Enable} we will show how these encodings can
        be used to prove properties of the Enable operator.

@@

\begin{definition}[Open automaton]
	\label{def:open-automaton}
	An \emph{open automaton} is a structure\\ $A =
	\langle J,\mathcal{S},s_0,\mathcal{T}\rangle$ where:
	\begin{itemize}
		\item[$\bullet$]  $I$ and $J$ are  two sets of indices,
		%	\item[$\bullet$]  $LTS_i^{i\in I}$ is a family of LTSs,
		\item[$\bullet$]  $\mathcal{S}$ is a set of states and $s_0 \in \mathcal{S}$ the initial state,
		\item[$\bullet$] $\mathcal{T}$ is a set of open transitions and, for each
		$t\in \mathcal{T}$, there exist $I'$, $J'$ with $I'\subseteq I$, $J'
		\subseteq J$, such that $t$ is an open transition over $J'$
		and  $\mathcal{S}$.

@@

\paragraph{States of open pNets.}\label{def-states}
A state of an open pNet is a tuple of the
states of its leaves (in which we denote tuples
in structured states as $\triangleleft\ldots\triangleright$).
For any pNet $p$, let $\overline{\mathit{Leaves}} = \mylangle S_i,{s_i}_0, \to_i\myrangle^{i \in L}$ be the set of PLTSs at its leaves,
then $\mathit{States}(p) = \{\ostate{s_i^{i\in L}}
|\, \forall i\in L. s_i \in S_i\}$.
A PLTS could be its own single leave:
$\mathit{States}(\mylangle S,s_0, \to\myrangle) = \{\ostate{s}|\, s \in S\}$.
The initial state is defined as:
$\mathit{InitState}(p) = \ostate{{{s_i}_0}^{i\in L}}$.

@@

\paragraph{Predicates.}
Let
$\mylangle\overline{\pNet},J,\symb{SV}_k^{k\in K} \myrangle$
be a pNet. Consider a synchronisation vector $SV_k$, for $k\in K$. We build a
predicate $\MkPred$ relating
the actions of the involved sub-pNets and the resulting actions. This predicate verifies:
\begin{multline*}
	\MkPred(SV_k, a_i^{i\in I}, b_j^{j\in J}, v)\Longleftrightarrow
	%% \begin{array}{l}
	\exists {(a'_i)}^{i\in I},
	{(b'_j)}^{j\in J},v'.\\
	SV_k={(a'_i)}^{i\in I}, {(b'_j)}^{j\in J}\rightarrow v'
	\land
	\forall i\in I.\, a_i=a'_i\land \forall j \in J.\, b_j=b'_j \land v=v'
\end{multline*}

@@

If we use the data-oriented pNet $EnableData$ from Figure
\ref{schema:enable-pnets} to build a similar $EnableDataCompLeft$
``P>>(Q>>R)'' expression, the corresponding open transition would be:

  \label{OT:example-2}
        \begin{mathpar}
	  OT_2'  =
          \inferrule*[myfraction=\reddottedrule]
	{\{\xrightarrow{\delta(x2)}_P ~~
          \xrightarrow{acc(x2)}_Q \} ~~
          [v_1=0 \land v_2 = 0 \land \forall z. acc(x2) \neq
            \delta(z)] ~~ \{v1:=1\}
        }
	{\ostate{0,0} \xrightarrow {\underline{\delta(x2)}} \ostate{0,0}}
	\end{mathpar}
        where $v_1$ and $v_2$ are respectively the state variables of
        the first and second EnableData pNets. Indeed the global
        state space of this version is reduced to $\ostate{0,0}$, and
        the state change is replaced by the assignment ``$v_1:=1$''.

@@

In \cite{henrio:Forte2016} we also proved the following result, which
ensures the termination of our semantic construction algorithm.
\begin{theorem}[Finiteness]
  \label{Th:finiteness}
   Let $pnet=\mylangle \overline{\pNet},\overline{S}, \symb{SV}_k^{k\in K}\myrangle$ be an open pNet  with leaves $l_i^{i\in I}$ and holes $h_j^{j\in
 	J}$, where the sets $I$ and $J$ are finite. Suppose the synchronisation vectors of all pNets 
 included in  $pnet$   are finite, and 
 % $\forall i \in L.\, finite{(states(l_i))} \text{ and } l_i$ 
 $l_i $ has a finite number of state variables for each $i\in I$. Then the semantics of $pnet$ is an open automaton 
 %$\mathcal{T}$ 
 with finitely many states and transitions.
\end{theorem}
  Remark that all the elements of such pNets and open automata are symbolic, and they can represent many classes of unbounded systems.

@@

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require A pNet $P$ (can be a PLTS, but not a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
	\For {each $\mathit{OT} \in \mathit{OTs}$}
          \State Add $\mathit{OT}$ to $L$;
          \State Let $S'$ be the target state of $\mathit{OT}$
          Add $S'$ to the unexplored states if necessary:
	  \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
	  \EndFor
\EndWhile
\State $\mathit{L1}$=filterUnsatTransitions($\mathit{L,Assignments}$);
\State $\mathit{OA_0}$=makeReachableSubAutomaton($\mathit{InitState}(P),L1)$;
\Repeat $\mathit{OA_n}$=RefineReachableSubAutomaton($\mathit{AO_{n-1}}$)
\Until  $\mathit{OA_n}=\mathit{OA_{n-1}}$; \Comment Fixpoint reached
\State \Return $\mathit{OA_n}$;

@@

Algorithm~\ref{alg1} starts with an open pNet, and builds its set of open
transitions. Its main loop is a classical residual algorithm: starting
from the initial global state, it picks a state in an unexplored set, 
computes all possible OTs, and adds the target states in the
unexplored set, until this set is empty.

Before the SAT checking step, it may contain symbolic transitions that
do not have valid ground instanciations. This is not logically
incorrect, but is naturaly not an optimal representation. So we prune
these UNSAT transitions: at the end of the open automaton generation,
the predicate of each OT is translated 
into SMTlib assertions (see Section~\ref{section:pruning}), and checked for
satisfiability (function \texttt{filterUnsatTransitions}). 
Then we compute the set of reachable transitions and states to build
the final open automaton (more details in section~\ref{section:BuildingTheOpenAutomaton}).

@@

\paragraph{Filtering.}
While matching a vector with a combination tuple, \emph{Matching}
tries to filter out some incompatibilities; there may be several
reasons why the matching would fail: 
\begin{itemize}
  \item if some subnet is marked as inactive in the vector, and the
    chosen combination has an active behaviour at this position, or vice-versa;
  \item if some subnet action expression in the vector does not match (by
    pattern-matching) the corresponding action expression in $C$;
  \item or if the whole set of active subnet actions in the vector
    cannot be matched (by unification) with the corresponding action expressions
      on the tuple $C$ (this is strictly stronger than the
      previous filter). 
\end {itemize}

@@

In a PLTS, there may be several incoming transitions 
that assign potentially different (symbolic) values to a state variable.
To handle such cases, the algorithm manages a
list of assignment expressions for each PLTS state. 
The only assignments for state variables that happen at the PLTS level
are as follows: every time the MakeTransitions 
algorithm is applied on a PLTS, it collects the expressions from the
$Post$ in each transition. 
% As there is no assignments collecting work on a pNet node, it is not
% shown in Algorithm~\ref{alg2}.

@@

It shows the case where processes $P$ and $R$ ``try'' to communicate directly, $P$, but this is made impossible by the combination of chosen synchronisation vectors and their guards. This
  mismatch is materialised by the predicate
fragment ``$a1 = \delta(x2) \land \forall z. a1 \neq \delta(z)$''. In Section~\ref{section:TranslationToSMTlib} we will show how this is encoded and checked using the SMT engine.

@@

A further reduction may now be gained by using the information from
these sets of possible assignments: adding this information when
checking satisfiability of transitions leaving a given state may
result in more Unsat transitions. The set of possible
assignments computed by the algorithm is an over approximation of
possible values of the 
variables of such state, so narrowing it means we get a better
approximation, and potentially more information, which can be used in
the SAT checking of subsequent transitions. In turn, this may reduce the possible
assignments of the corresponding target states. In
algorithm~\ref{alg1}, we iterate this procedure, each iteration
reducing the set of 
transitions and the sets of assignments; the final open automaton
is the fixed point of this iteration.

In practice, this could be
expensive, and it is not incorrect to leave some Unsat OTs in the
result. From a  practical point of view, very little will be gained to compute the
fixed point, so we would only perform one step of iteration. Further algorithms 
(e.g. model-checking or equivalence checking) will be able to use the
full information from the predicates and the assignment sets.

@@

The basic algorithm from the previous section potentially explores some
areas of the global state-space that will not be reachable after
checking the OTs' predicates for satisfiability. Incorporating the
satisfiability check inside the residual algorithm will definitely
save some of these useless computation. More precisely, it will still
create, and submit to Z3, the OTs at the ``border'' of the state-space, but
not the transitions issued from non-reachable global states. We use
this idea to create our ``smart'' algorithm.

>>>>

Implementing these semantics raised several challenges:
\begin{itemize}
	\item to obtain a tool that could be applied to pNets representing
	various languages, in particular various action algebras,
	with their specific decision theories;
	\item to separate clearly the 
	algorithm generating the transitions of an open automaton from
	combination of all possible (symbolic) behaviours, and from
	the symbolic reasoning part, specifically here using an SMT
	engine to check the
	satisfiability of the predicates generated by our algorithm;
	\item to build a prototype and validate the approach on our basic
	case-studies, and understand the efficiency of the interaction
	with the SMT solver.
\end{itemize}

@@

\paragraph{Contribution.}
In the current work we contribute in the following aspects.
\begin{itemize}
	\item We present an open automaton generation algorithm. We
	implement a full working prototype, within the 
	VerCors platform. In the process, we improve the
	semantic rules from~\cite{henrio:Forte2016}, and add features in
	the algorithm to deal with the full 
	model, including management of variables and assignments.
	\item We implement the interaction between our algorithm and the Z3
	SMT solver, for checking the satisfiability of the transitions
	generated by the algorithm.
	\item We define a strategy checking satisfiability of open transitions in the course of the residual algorithm, and show that  the reachable part of the open automaton can be generated and it is minimal with respect to the decidability of predicate satisfiability.
	\item We show the usage of this approach on various use-cases from different settings, including an
	industry-inspired case-study, namely one architectural pattern
	extracted (and extended) from the BIP specification of a nanosatellite on-board
	software.
\end{itemize}

@@

\paragraph{Structure.}
In Section
\ref{section:pnets} we give a description and a formal definition of
the pNet model. 
In Section \ref{section:op-semantics} we briefly recall  the operational semantics
of pNets.
In Section \ref{section:implementation} we present an algorithm to compute this semantics, including the interaction with Z3.
In Section~\ref{section:use-cases} we give a few use-cases, in particular, a BIP
architecture from the nanosatellite case-study. 
Finally we conclude and discuss perspectives in Section
\ref{section:conclusion}.

@@

\renewcommand{\P}{\mathcal V}
\def\Talg{\mathcal{T}_{\Sigma,\P}}
Formally, we assume the existence of a term algebra $\Talg$,
where $\Sigma$ is the signature of the data and action constructors,
and $\P$ a set of variables. Within $\Talg$, we distinguish a set of
data expressions $\mathcal{E}_\P$, including a set of Boolean
expressions $\mathcal{B}_{\P}$ ($\mathcal{B}_{\P}\subseteq\mathcal{E}_\P$).
On top of $\mathcal{E}_\P$ we build the action algebra
$\mathcal{A}_\P$, with $\mathcal{A}_\P\subseteq\mathcal{T}_\P,
\mathcal{E}_\P\cap\mathcal{A}_\P=\emptyset$;
 action terms will use data expressions as sub-terms.
The function $\vars(t)$ identifies the set of variables in a term
$t\in\AlgT$.

@@

\begin{definition}[PLTS]
	\label{PLTS}
	%\noteSB{It's a bit strange that algebra presentations are not referred to here}
	Given a term algebra $\Talg$, a PLTS is a tuple
	$\langle S,s_0, \to\rangle$, where
	\begin{itemize}
		\item[$\bullet$]
		$S$ is a set of \emph{states}, with $s_0 \in S$ the \emph{initial state}.
		\item[$\bullet$] $\to \subseteq S \times L \times S$ is the \emph{transition relation}, with
		$L$ the set of labels of the form
		$\langle \alpha,~e_b,~(x_j\!:= {e}_j)^{j\in J}\rangle$,
		where $\alpha \in\AlgA_{\P}$ is a parameterised action,
		$e_b \in \AlgB_{\P}$ is a guard, and
		expressions  $\AlgE_{\P}\cup\AlgA_{P}$ are assigned to $x_j$.
		If 
		$s \xrightarrow{\langle \alpha,~e_b,~(x_j\!:= {e}_j)^{j\in
				J}\rangle} s'$ then 
		$\vars(e_b)\!\subseteq\! \vars(s)\cup\vars(\alpha)$, and
		$\forall j\!\in\! J .\,\vars(e_j)\!\subseteq\! \vars(s)\land 
		x_j\!\in\!\vars(s')$.
		%\item[$\bullet$]
		%The $x_j$ are \emph{state variables} of %% the state $s'$. 
	\end{itemize}
\end{definition}

@@

%% \begin{definition}[pNets]\label{def-pnets}
%% A pNet is a hierarchical structure, whereof the leaves are PLTSs and holes:
%% $\pNet::= PLTS~|~\langle \pNet_i^{i\in I}, S_j^{j\in J}, \symb{SV}_k^{k\in K}\rangle$,
%% where
%% \begin{itemize}
%% \item[$\bullet$] %\noteEM{Pire que ca, $\I$ n'est pas
%%                  %defini...}\noteSB{Why is one $\I$ and the other
%%   %$\I_\P$?}
%%   $I$ and $J$ are the sets over which are indexed sub-pNets and holes, respectively, such that: $I\cap J=\emptyset$ and  $I\cup J\neq\emptyset$.

@@

\begin{definition}[pNets]\label{def-pnets}
	A pNet is a hierarchical structure where leaves are PLTSs and holes:
	%%  \footnote{Typing can 
	%% be added to pNets by defining and checking their interfaces, called 
	%% sort~\cite{HMZ-FORTE2016}}
	$\pNet\triangleq PLTS~|~\langle \pNet_i^{i\in I}, J, \symb{SV}_k^{k\in 
		K}\rangle$, with $I, J, K$ potentialy infinite,
	where
	\begin{itemize}
		\item[$\bullet$] $\pNet_i^{i\in I}$ is the family of sub-pNets.
		%  $\pNet_i^{i\in I}$ is a family of sub-pNets where $I\in\I_\P$ is the set over which 
		%sub-pNets are indexed.

		\item[$\bullet$] $J$ is a set of indexes, called \emph{holes}.
		$I$ and $J$ are \emph{disjoint}: $I\!\cap\! J=\emptyset$,  $I\!\cup\! J\neq\emptyset$.
		%\item[$\bullet$] $\Sort_j \subseteq \AlgAS$ is a set of action terms, denoting the 
		%\emph{sort} of
		%hole $j$.

		\item[$\bullet$] $\symb{SV}_k^{k\in K}$ is a set of
		synchronisation vectors ($K\in\I_\P$).  It is required that \linebreak $\forall k\!\in\! K,
		\symb{SV}_k\!=\!\alpha_{l}^{l\in I_k \uplus J_k}\to\alpha'_k\,|\,g_k$, where
		$\alpha'_k\in \actions{\variables}$, $I_k\subseteq I$, $J_k\subseteq J$, and 
		$\vars(\alpha'_k)\subseteq \bigcup_{l\in I_k\uplus 
			J_k}{\vars({\alpha_l})}$. The global action of a vector $\symb{SV}_k$ is
		$\alpha'_k$. The Boolean expression $g_k $, with $\vars(g_k)\subseteq \bigcup_{l\in 
			I_k\uplus J_k}{\vars({\alpha_l})}$, is a guard associated to the vector.

@@

  \includegraphics[width=\linewidth]{ActaXFIG/Enable1}
  \\[1.3ex]
 \includegraphics[width=\linewidth]{ActaXFIG/Enable2}
  \caption{Two pNet encodings for  Enable }  \label{schema:enable-pnets}
\end{minipage}
  \hspace{2mm}
\begin{minipage}{6cm}
  \includegraphics[width=\linewidth]{ActaXFIG/P-QR}
  \caption{Composed pNet for ``$P>>(Q>>R)$''}  \label{schema:enable-composed}
\end{minipage}

@@

	To give an intuition of the open pNet model and its semantics, we use
	here a simple example based on the ``Enable'' operator of the LOTOS language.
	In the Enable expression ``$P>>Q$'', an \texttt{exit} statement within $P$
	terminates the current process, carrying a value $x$ that
	is captured by the \texttt{accept} statement of $Q$. 	
	We use a simple action algebra,
	containing two constructors $\delta(x)$ and $acc(x)$, for any
	possible data type of the variable $x$, corresponding to the
	statements \texttt{exit} and \texttt{accept}. 
	We need no specific predicate over the action expressions, apart
	from equality of actions.

@@

	Note that synchronisation vectors are defined in a parameterised
	manner:  the first and third lines represent one vector
	for each possible action of hole $P$
	(resp. $Q$). This notation can also use predicates, as in the first
	case, in which we want the vector to apply to any action of $P$
	except for $\delta(z)$.

	In Figure \ref{schema:enable-composed}, we enrich our example by
	composing two occurences of the $EnableState$ pNet. In Section
        \ref{section:use-case:Enable} we will show how these encodings can
        be used to prove properties of the Enable operator.

@@

\begin{definition}[Open automaton]
	\label{def:open-automaton}
	An \emph{open automaton} is a tuple $
	\langle I, J,\mathcal{S},s_0,\mathcal{T}\rangle$, where
	\begin{itemize}
		\item[$\bullet$]  $I$ and $J$ are  two sets of indices,
		%	\item[$\bullet$]  $LTS_i^{i\in I}$ is a family of LTSs,
		\item[$\bullet$]  $\mathcal{S}$ is a set of states and $s_0 \in \mathcal{S}$ the initial state,
		\item[$\bullet$] $\mathcal{T}$ is a set of open transitions and, for each
		$t\in \mathcal{T}$, there exist $I'$, $J'$ with $I'\subseteq I$, $J'
		\subseteq J$, such that $t$ is an open transition over $J'$
		and  $\mathcal{S}$.

@@

\paragraph{States of open pNets.}\label{def-states}
A state of an open pNet is a tuple of the
states of its leaves (in which we denote tuples
in structured states as $\triangleleft\ldots\triangleright$).
For any pNet $p$, let $\overline{\mathit{Leaves}} = \langle S_i,{s_i}_0, \to_i\rangle^{i \in L}$ be the set of PLTSs at its leaves,
then $\mathit{States}(p) = \{\ostate{s_i^{i\in L}}
|\, \forall i\in L. s_i \in S_i\}$.
A PLTS could be its own single leave:
$\mathit{States}(\langle S,s_0, \to\rangle) = \{\ostate{s}|\, s \in S\}$.
The initial state is defined as:
$\mathit{InitState}(p) = \ostate{{{s_i}_0}^{i\in L}}$.

@@

\paragraph{Predicates.}
Let
$\langle\overline{\pNet},J,\symb{SV}_k^{k\in K} \rangle$
be a pNet. Consider a synchronisation vector $SV_k$, for $k\in K$. We build a
predicate $\MkPred$ relating
the actions of the involved sub-pNets and the resulting actions. This predicate verifies:
\begin{multline*}
	\MkPred(SV_k, a_i^{i\in I}, b_j^{j\in J}, v)\Longleftrightarrow
	%% \begin{array}{l}
	\exists {(a'_i)}^{i\in I},
	{(b'_j)}^{j\in J},v'.\\
	SV_k={(a'_i)}^{i\in I}, {(b'_j)}^{j\in J}\rightarrow v'
	\land
	\forall i\in I.\, a_i=a'_i\land \forall j \in J.\, b_j=b'_j \land v=v'
\end{multline*}

@@

If we use the data-oriented pNet $EnableData$ from Figure
\ref{schema:enable-pnets} to build a similar $EnableDataCompLeft$  expression
``$P>>(Q>>R)$'', the corresponding open transition would be:

  \label{OT:example-2}
        \begin{mathpar}
	  OT_2'  =
          \inferrule*[myfraction=\reddottedrule]
	{\{\xrightarrow{\delta(x2)}_P ~~
          \xrightarrow{acc(x2)}_Q \} ~~
          [v_1=0 \land v_2 = 0 \land \forall z. acc(x2) \neq
            \delta(z)] ~~ \{v1:=1\}
        }
	{\ostate{0,0} \xrightarrow {\underline{\delta(x2)}} \ostate{0,0}}
	\end{mathpar}
        where $v_1$ and $v_2$ are respectively the state variables of
        the first and second $EnableData$ pNets. Indeed the global
        state space of this version is reduced to $\ostate{0,0}$, and
        the state change is replaced by the assignment ``$v_1:=1$''.

@@

In \cite{henrio:Forte2016} we also proved the following result, which
ensures the termination of our semantic construction algorithm.
\begin{theorem}[Finiteness]
  \label{Th:finiteness}
   Let $pnet=\langle \overline{\pNet},\overline{S}, \symb{SV}_k^{k\in K}\rangle$ be an open pNet  with leaves $l_i^{i\in I}$ and holes $h_j^{j\in
 	J}$, where the sets $I$ and $J$ are finite. Suppose the synchronisation vectors of all pNets 
 included in  $pnet$   are finite, and 
 % $\forall i \in L.\, finite{(states(l_i))} \text{ and } l_i$ 
 $l_i $ has a finite number of state variables for each $i\in I$. Then the semantics of $pnet$ is an open automaton 
 %$\mathcal{T}$ 
 with finitely many states and transitions.
\end{theorem}
  Remark that all the elements of such pNets and open automata are symbolic, and they can represent many classes of unbounded systems.

@@

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require A pNet $P$ (can be a PLTS, but not a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
	\For {each $\mathit{OT} \in \mathit{OTs}$}
          \State Add $\mathit{OT}$ to $L$;
          \State Let $S'$ be the target state of $\mathit{OT}$;
          add $S'$ to the unexplored states if necessary:
	  \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
	  \EndFor
\EndWhile
\State $\mathit{L1}$=filterUnsatTransitions($\mathit{L,Assignments}$);
\State $\mathit{OA_0}$=makeReachableSubAutomaton($\mathit{InitState}(P),L1)$;
\Repeat $\mathit{OA_n}$=RefineReachableSubAutomaton($\mathit{OA_{n-1}}$)
\Until  $\mathit{OA_n}=\mathit{OA_{n-1}}$; \Comment Fixpoint reached
\State \Return $\mathit{OA_n}$;

@@

Algorithm~\ref{alg1} starts with an open pNet and builds its set of open
transitions. Its main loop is a classical residual algorithm: starting
from the initial global state, it picks a state in an unexplored set, 
computes all possible OTs, and adds the target states in the
unexplored set, until this set is empty.

Before the SAT checking step, it may contain symbolic transitions that
do not have valid ground instanciations. This is not logically
incorrect, but is naturally not an optimal representation. So we prune
these UNSAT transitions: at the end of the open automaton generation,
the predicate of each OT is translated 
into SMTlib assertions (see Section~\ref{section:pruning}), and checked for
satisfiability (function \texttt{filterUnsatTransitions}). 
Then we compute the set of reachable transitions and states to build
the final open automaton (more details in Section~\ref{section:BuildingTheOpenAutomaton}).

@@

\paragraph{Filtering.}
While matching a vector with a combination tuple, \emph{Matching}
tries to filter out some incompatibilities; there may be several
reasons why the matching would fail: 
\begin{itemize}
  \item if some subnet is marked as inactive in the vector, and the
    chosen combination has an active behaviour at this position, or vice-versa;
  \item if some subnet action expression in the vector does not match (by
    pattern-matching) the corresponding action expression in $C$;
  \item or if the whole set of active subnet actions in the vector
    cannot be matched (by unification) with the corresponding action expressions
      in the tuple $C$ (this is strictly stronger than the
      previous filter). 
\end {itemize}

@@

In a PLTS, there may be several incoming transitions 
that assign potentially different (symbolic) values to a state variable.
To handle such cases, the algorithm manages a
list of assignment expressions for each PLTS state. 
The only assignments for state variables that happen at the PLTS level
are as follows: every time the $MakeTransitions$ method 
 is applied on a PLTS, it collects the expressions from the
$Post$ in each transition. 
% As there is no assignments collecting work on a pNet node, it is not
% shown in Algorithm~\ref{alg2}.

@@

It shows the case where processes $P$ and $R$ ``try'' to communicate directly, but this is made impossible by the combination of chosen synchronisation vectors and their guards. This
  mismatch is materialised by the predicate
fragment ``$a1 = \delta(x2) \land \forall z. a1 \neq \delta(z)$''. In Section~\ref{section:TranslationToSMTlib} we will show how this is encoded and checked using the SMT engine.

@@

A further reduction may now be gained by using the information from
these sets of possible assignments: adding this information when
checking satisfiability of transitions leaving a given state may
result in more UNSAT transitions. The set of possible
assignments computed by the algorithm is an over approximation of
possible values of the 
variables of such state, so narrowing it means we get a better
approximation, and potentially more information, which can be used in
the SAT checking of subsequent transitions. In turn, this may reduce the possible
assignments of the corresponding target states. In
Algorithm~\ref{alg1}, we iterate this procedure, each iteration
reducing the set of 
transitions and the sets of assignments; the final open automaton
is the fixed point of this iteration.

In practice, this could be
expensive, and it is not incorrect to leave some UNSAT OTs in the
result. From a  practical point of view, very little will be gained to compute the
fixed point, so we would only perform one step of iteration. Further algorithms 
(e.g. model-checking or equivalence checking) will be able to use the
full information from the predicates and the assignment sets.

@@

The basic algorithm from the previous section potentially explores some
areas of the global state space that will not be reachable after
checking the OTs' predicates for satisfiability. Incorporating the
satisfiability check inside the residual algorithm will definitely
save some of these useless computation. More precisely, it will still
create, and submit to Z3, the OTs at the ``border'' of the state-space, but
not the transitions issued from non-reachable global states. We use
this idea to create our ``smart'' algorithm.
